---
title: "lattice legends"
author: "Steve Pittard"
date: "February 11, 2016"
output: pdf_document
---

In class on Wed I told you about a little problem with lattice graphics and how it creates an automatic legend that can have different plot characters than what you had specified in the graph. As an example let's begin with the following:

```{r}
library(lattice)
xyplot(mpg~wt,data=mtcars,
              group=cyl,
              auto.key=list(columns=3))

```

Okay next we'll pick a different plot character such as 19 which is a filled point. However, the resulting legend doesn't have the filled points. 

```{r}
library(lattice)
xyplot(mpg~wt,data=mtcars,
              group=cyl,
              auto.key=list(columns=3),
              pch=19)

```

In class I showed you a cople of workarounds but I thought I would show you another that might be easier although it is more verbose - that is there is more to type. In this case
we include the **par.settings** argument which takes a list as an argument. Within that list we can specify any number of parameters that impact the look of the legend such as the
plot character (pch), size (cex), and the auto.key

```{r}
library(lattice)
xyplot(mpg~wt,data=mtcars,
              group=cyl,
              type=c("p","g"),
              main="MPG vs Weight",
              par.settings=list(superpose.symbol=list(pch=c(19), cex=1.5)), 
                                                       auto.key=list(space="right"))
                                                     
                                                    
```


```{r}
library(lattice)
xyplot(mpg~wt,data=mtcars,
              group=cyl,
              type=c("p","g"),
              main="MPG vs Weight",
              par.settings=list(superpose.symbol=list(pch=c(18), cex=1.5)), 
                                                      auto.key=list(space="right"))
                                                     
                                                    
```


You could even pass in some colors if you wanted something other than the default colors provided by lattice. 


```{r}
library(lattice)
colors <- rainbow(3)

xyplot(mpg~wt,data=mtcars,
              group=cyl,
              type=c("p","g"),
              main="MPG vs Weight",
              col=colors,
              par.settings=list(superpose.symbol=list(col=colors,pch=c(19), cex=1.5)),
                                                       auto.key=list(space="right"))
                                                     
```

The only weakness of this approach is that the we are passing in a list within a list so it is easy to type too many parentheses (or not enough). As I mentioned in class this kind of thing can become quite tedious. Now, to generalize this kind of thing you should probably  create a function out of this and pass in arguments such as pch and colors to generalize this. Just substitute in the arguemnt names into the par.settings list above. 