BIOS545 Week 3 Lab 2: Data Frames and Control Structures
============================================================
Pittard wsp@emory.edu


## DataFrames Part 2

Practice 1.0) Merging Data Frames
```{r eval=FALSE}
df1  <- data.frame(indiv_id = 1:4, snp1 = c(1,1,0,1), snp2 = c(1,1,0,0)) 

df2 <- data.frame(indiv_id = c(1,3,4,6), cov1 = c(1.14,4.50,0.80,1.39), 
                                          cov2 = c(74.6,79.4,48.2,68.1))

merge(tb1, tb2, by="indiv_id", all=TRUE)  # Brings in all data from both data frames

# Excludes indiv_id rows from tb2 that don't also appear in tb1

merge(tb1, tb2, by="indiv_id", all.x=T)   

# Excludes indiv_id rows from tb1 that don't also appear in tb2

merge(tb1, tb2, by="indiv_id", all.y=T)

# Let's change the names in tb2

names(tb2) <- c("id","cov1","cov2")

# Let's do a merge wherein we specify "keys" from tb1 and tb2

merge(tb1,tb2,by.x="indiv_id",by.y="id",all=TRUE)
```

***

1.1) Let's use the split function on mtcars to split it up by cylinder groups

```{r eval=FALSE}

# This will create a list, (hold), wherein each element is a data frame. 

hold <- split(mtcars, mtcars$cyl) 

# The name of each element comes from the unique values taken by the "cyl" variable

names(hold)

# Once we have things stashed in the list we can use sapply (or lapply) to 
# loop over each element and apply some function 

sapply(hold,nrow)

hold <- split(mtcars,mtcars$cyl) 

sapply(hold, function(x) mean(x$mpg))

```

***

1.2) We can also sort a data frame although we can't use the sort command directly as we would with a single vectors.

```{r eval=FALSE}

# We use the order function to get back row numbers in the order of MPG from lowest to highest

order(mtcars$mpg) # Returns row numbers representing lowest to highest MPG

mtcars[15,]  # The Cadillac has the worst gas mileage

mtcars[20,]  # The Corolla has the best gas mileage

head(newdata)

newdata <- mtcars[order(mtcars$mpg),] # Sortson MPG from low to hi

head(newdata)

newdata <- mtcars[rev(order(mtcars$mpg)),]  # Sorts on MPG from hi to lo

head(newdata)

newdata <- mtcars[order(-mtcars$mpg),] # Same as above

# We can also sort by two or more keys. Like first by cylinder then MPG

newdata <- mtcars[order(mtcars$cyl,mtcars$mpg),] # Cyl: low to hi, MPG low to hi

newdata <- mtcars[order(mtcars$cyl,-mtcars$mpg),] # Cyl: low to hi, MPG hi to lo


```

***

1.3) Sampling from a dataframe can be done also. We have to figure out the
number of rows and use that as input to the sample function

```{r eval=FALSE}

my_records <- sample(1:nrow(mtcars), 10, replace = FALSE)

my_records  # Random row numbers from mtcars

sample_of_ten <- mtcars[my_records,] # Use my_records as input to brackets

sample_of_ten

```

***

## Control Structures

2.0) for loops are for processing elements of vectors, matrices, and lists 

```{r eval=FALSE}

# for loops are simple

for (ii in 1:3) {
  print(ii)
}

x <- rnorm(3)
for (ii in 1:length(x)) {    # We start with the last element number
   print(x[ii])
}

# Here we will loop over some x values to get some y values to plot

y <- vector()  # A blank vector

x <- 1:6

for (ii in 1:length(x)) {
  y[ii] <- x[ii]^2
}

plot(x,y,main="Super Cool Data Plot",type="b",pch=19,col="red")


```

***

2.1) We can use for loops to do calculations

```{r eval=FALSE}

set.seed(188)

x <- rnorm(1000)  # 1,000 random elements from a N(20,4)

# Set mymin to the first value in x. Then use the for loop to 
# compare it to all other values in x. If we find a value that is less
# than the current mymin then we set mymin to it.

mymin <- x[1] 

for (ii in 1:length(x)) {
  if (x[ii] < mymin) {
     mymin <- x[ii]
  }
}

mymin

```

***

2.2) This approach can be applied to processing data frames like mtcars

```{r eval=FALSE}

mpgsum = 0

for (ii in 1:nrow(mtcars)) {
  mpgsum <- mpgsum + mtcars[ii,"mpg"]
}

mpgmean <- mpgsum/nrow(mtcars)   # Divide the sum by the # of records

cat("Mean MPG for all cars is:",mpgmean,"\n")

```

***

2.3) We can use the for loop in conjunction with the split function to process list elements

```{r eval=FALSE}

mysplit <- split(mtcars,mtcars$cyl)

for (ii in 1:length(mysplit)) {
   print(nrow(mysplit[[ii]]))
}

# This is equivalent to doing the following

sapply(mysplit, nrow)

```

***

2.4) We can access the list element names and values as we loop

```{r eval=FALSE}

mysplit <- split(mtcars,mtcars$cyl)

for (ii in 1:length(mysplit)) {
   splitname <- names(mysplit[ii])
   cat("mean for",splitname,"cylinders is",mean(mysplit[[ii]]$mpg),"\n")
}

# This is basically equivalent to 

sapply(mysplit, function(x) mean(x$mpg))


```

***

2.5) Looping over matrices is easy with for loops. However we have to use two of them. One for rows and one for columns

```{r eval=FALSE}
set.seed(123)

mymat <- matrix(round(rnorm(6),2),3,2)

for (ii in 1:nrow(mymat)) {
    cat("The sum of row",ii,"is",sum(mymat[ii,]),"\n")
}

# The sum of row 1 is -0.49 
# The sum of row 2 is -0.1 
# The sum of row 3 is 3.28 

# This will generate the sum of all columns

for (ii in 1:ncol(mymat)) {
    cat("The sum of col",ii,"is",sum(mymat[,ii]),"\n")
}

# The sum of col 1 is 0.77 
# The sum of col 2 is 1.92 

```

***

### if statements

3.0) If statements allow us to make decisions based on the value of a vector, list, or matrix element.

```{r eval=FALSE}

some.num <- 3               

if (some.num < 3) {        # A more involved if statement
     print("Less than 3")
} else if (some.num > 3) {
     print("Greater than 3")
} else {
     print("Must be equal to 3")
}

# if statements are frequently used for error checking

x <- 4 ; y <- 5

if (!is.numeric(x) | !is.numeric(y)) {
   stop("I need numeric values to do this")
} else {
   if (x == y) {
       print("Equal")
   } else {
      print("Not equal")
   }
}
```

***

3.1) the ifelse statement is useful for processing vectors

```{r eval=FALSE}
some.data <- rnorm(10000,0,2)

colors <- ifelse(some.data < 0,"RED","GREEN")

plot(some.data,col=colors)

# 

mtcars$rating <- ifelse(mtcars$mpg >= mean(mtcars$mpg), "blue", "red")

plot(mtcars$mpg~mtcars$wt,col=mtcars$rating,pch=19, main="MPG vs wt")

grid()

legend("topright", c("> mean","< mean"), pch=19,
        col=c("blue","red"),title="Legend",cex=0.7)

```

***

3.2) for and if loops are frequently found together in the same structure

```{r eval=FALSE}

gradevec <- vector()  # Initialize an empty vector to containt the grades

score <- c(74,68,98,90,100,67,59)

for (ii in 1:length(score)) {
  if (score[ii] == 100) {
      gradevec[ii] <- "A+"
  } else if (score[ii] >= 90 & score[ii] <= 99) {
      gradevec[ii] <- "A"
  } else if (score[ii] >= 80 & score[ii] <= 89) {
      gradevec[ii] <- "B"
  } else if (score[ii] >= 70 & score[ii] <= 79) {
      gradevec[ii] <- "C"
  } else if (score[ii] >= 60 & score[ii] <= 69) {
      gradevec[ii] <- "D"
  }
  else {
    gradevec[ii] <- "F"
  }
}
gradevec

```

***

3.3) We can use the for loop / if statement approach to process vectors

```{r eval=FALSE}
set.seed(123)

x <- round(runif(10,1,20))

for (ii in 1:length(x)) {
    if (x[ii] %% 2 == 0) {
        print(TRUE)
    }
    else {
        print(FALSE)
    }
}

# We can mimic the bracket notation by processing elements and storing computations in a vector

set.seed(123)

x <- round(runif(10,1,20))

logvec <- vector()            # Setup an empty vector

for (ii in 1:length(x)) {
    if (x[ii] %% 2 == 0) {
        logvec[ii] <- TRUE
    }
    else {
        logvec[ii] <- FALSE
    }
}

logvec

x[logvec]


```

***

### while loops

3.4) The while loop exists when you might not know in advance how many times you need to iterate

```{r eval=FALSE}
sum <- 0 
n <- 1000
i <- 1
while (i <= n) {
    sum <- sum + i
    i <- i + 1
}

sum

# Example of convergence. Take the square root of a number, say 13, and then
# take the square root of that , and so on. This will converge to 1. Here is
# how we could this with a while loop.

num <- 13
sqrtval <- sqrt(num)

# Loop until the sqrt value becomes equal to 1

while ( sqrtval != 1)  {    
  sqrtval <- sqrt(sqrtval)

# sprintf allows us to format a variable according to a pattern
# See http://www.cookbook-r.com/Strings/Creating_strings_from_variables/
# We want to see the value of sqrtval to 12 digits of precision

  print(sprintf("%2.12f",sqrtval))
}

```

***