BIOS 545 Spring 2015 Week 5 Lab 2: Graphics Part 2
============================================================
Pittard wsp@emory.edu
***

## Base Graphics Cont - Lattice graphics

PRACTICE 1.0) LET'S DRAW SOME BARPLOTS USING COLORS AND ALSO LOOK AT SOME "COLOR RAMPS"

```{r eval=TRUE}
myt <- tapply(mtcars$mpg, mtcars$cyl, mean)  

# tapply produces a table

barplot(myt, main = "Mean MPG per Cylinder Group", col=heat.colors(3))

myt <- tapply(mtcars$mpg, mtcars$carb, mean)

barplot(myt, main = "Mean MPG per Carb group", 
             col= terrain.colors(length(myt)),legend=T)

barplot(rev(sort(myt)), main = "Mean MPG per Carb group", 
            col= terrain.colors(length(myt)),legend=T)

mycols <- colorRampPalette(c("red","blue"))( 6 ) 


barplot(rev(sort(myt)), 
        main = "Mean MPG per Carb group", col=mycols, legend=T)


myt <- tapply(mtcars$mpg, mtcars$cyl, mean)  

# tapply produces a table

barplot(rev(sort(myt)), main = "Mean MPG per Cylinder Group",   col=heat.colors(3))

```
***

## More Annotation
PRACTICE 1.2) WE CAN ALSO ANNOTATE OUR BARPLOTS

```{r eval=TRUE}
myt <- tapply(mtcars$mpg, mtcars$cyl, mean)  

# tapply produces a table

temp <- barplot(myt, main = "Mean MPG per Cylinder Group",
                    col=heat.colors(3))

temp        # These represent the X coordinates for each bar

#     x    y   text

text(0.7, 10, 26.66)    # We put up the means one by one

text(1.9, 10, 19.74)

text(3.1, 10, 15.10)

# This will put up the text in one go


temp <- barplot(myt, main = "Mean MPG per Cylinder Group",
                    col=heat.colors(3))

text(temp,10,round(myt,2))
```
***

## Boxplots

PRACTICE 1.3) HERE WE CAN DRAW THE BOXPLOT VALUES ON THE ACTUAL GRAPH

```{r eval=TRUE}
fivenum(mtcars$mpg) # RETURNS THE FIVE NUM SUMMARY

boxplot(mtcars$mpg, main="MPG Boxplot",xlab="MPG",col="aquamarine")

# LET'S PUT UP A BOXPLOT WITH TWO COLORS

boxplot(mpg~am, data = mtcars, main="MPG / Transmission",
                col=heat.colors(2), names=c("auto","manual"))

legend(0.5,33,c("Auto","Manual"),fill=heat.colors(2))

# THIS WILL AUTOMATICALLY GENERATE THREE BOXPLOTS - ONE FOR
# EACH CYLINDER GROUP

boxplot(mpg~cyl,data=mtcars,main="MPG by Cylinders",
        col=rainbow(3), horizontal=TRUE, 
        names=c("Four","Six","Eight"))

# THIS IS THE SAME AS ABOVE EXCEPT WE USE A COLOR PALETTE

boxplot(mpg~cyl, main="MPG vs. Cylinder", data=mtcars, 
         notch=TRUE, col=terrain.colors(3))

# THE BOXPLOT FUNCTION RETURNS INFORMATION WE CAN USE

boxplot(mtcars$mpg, main="MPG Boxplot",xlab="MPG",col="aquamarine")


# WE SAVE THE PLOT INTO A VARIABLE FOR LATER INSPECTION AND USE

hold <- boxplot(mtcars$mpg, main="MPG Boxplot",
               xlab="MPG",col="aquamarine")
hold

text(1,10.4,10.4)
text(1,15.35,15.35)
text(1,19.20,19.20)
text(1,22.8,22.8)
text(1,33.9,33.9)


# OR could put up text in one go

text(1,hold$stats[,1],hold$stats[,1])

tmpvar <- boxplot(mpg~cyl,data=mtcars)


tmpvar$stats

tmpvar$stats[3,]

mtext(at=1:3, text=tmpvar$stats[3,])

# THIS IS THE SAME AS DOING THE FOLLOWING:


mtext(at=1, text=tmpvar$stats[3,1], side=3) # top margin
mtext(at=2, text=tmpvar$stats[3,2], side=3) # top margin
mtext(at=3, text=tmpvar$stats[3,3], side=3) # top margin

```
***

## Lattice Graphics

PRACTICE 2.0) LATTICE GRAPHICS AFFORDS US SOME CONVENIENCE THAT THE BASE GRAPHICS DOES NOT. 
```{r eval=TRUE}
library(lattice)  # Required for all lattice plot commands

# We can specify attributes of the points and the graph using arguments

xyplot(mpg~wt|factor(carb), data=mtcars, pch=19,main="MPG vs Wt per Carb #", type=c("p","g"))

# Lattice graphics let's us take advantage of factors in a way that base graphics does not

mtcars$am <- factor(mtcars$am, labels=c("auto","manual"))

mtcars$am

levels(mtcars$am)

# We want to plot mpg vs wt as a function of cylinder groups. this is really easy 

xyplot(mpg ~ wt, data = mtcars, pch = 19, main = "MPG vs. Weight")

xyplot(mpg ~ wt, data = mtcars, pch = 19, main = "MPG vs. Weight", type=c("p","g"))

xyplot(mpg ~ wt | factor(cyl), data = mtcars, pch=16, main = "MPG vs. wt by Cylinder Group")

xyplot(mpg ~ wt | factor(cyl), data = mtcars, pch=16, type=c("p","g"), main = "MPG vs. wt by Cylinder Group", layout=c(1,3))

```
***

PRACTICE 2.1) LATTICE GRAPHICS / xyplot CONTINUED

```{r eval=TRUE}

# We can adjust the layout easily

library(lattice)

xyplot(mpg~wt|factor(cyl),data = mtcars,layout=c(3,1))

# the conveniences afforded by lattice graphics work with other lattice functions

histogram(~mpg, data = mtcars, main = "MPG for all groups")

histogram(~mpg | factor(cyl), data = mtcars, layout = c(3,1))

# here we can use the auto.key argument to create a legend automatically.

xyplot(mpg~wt, data=mtcars, groups=factor(am), type=c("p","g"),
       main="MPG vs. Wt", auto.key=TRUE)

# OK


xyplot(mpg~wt, data=mtcars, groups=cyl, main="MPG vs. Wt", auto.key=TRUE)


# the auto.key argument supports a list structure that allows for customization

xyplot(mpg~wt, data=mtcars, groups=factor(am), 
       main="MPG vs. Wt", 
       auto.key=list(title="Transmission Type",cex=0.6,columns=2))

# The legend can grow and grow

xyplot(mpg~wt, data=mtcars, groups=factor(am), main="MPG vs. Wt", pch=19,
       auto.key = list(title="Transmission Type",
                       cex=0.6,columns=2,space="right"))

# here we tell lattice where to put the legend explicitly

xyplot(mpg~wt, data=mtcars, groups=factor(am), main="MPG vs. Wt", 
       pch=19, type=c("p","g"),
       auto.key = list(title="Transmission Type", cex=0.6, columns=2,
       corner=c(1,1)))


xyplot(mpg ~ wt | factor(cyl), data=mtcars, 
        pch=16,layout=c(3,1), main="MPG vs Wt per Cylinder",
        type=c("p","g"))


xyplot(mpg ~ wt | factor(cyl), data=mtcars, 
       pch=16,layout=c(3,1),
       groups = factor(am,labels=c("Auto","Manual")),
       type=c("p","g"),auto.key = TRUE)

xyplot(mpg ~ wt | factor(cyl), data=mtcars, 
        main = "MPG vs Wt per cylinder Group",
        groups = factor(am), pch=21, type=c("p","g"), 
        layout = c(1,3), auto.key=list(columns=2))

```
***
