BIOS560R Spring 2014 - LAB Matrices, Factors, Lists
============================================================
Pittard wsp@emory.edu
***

## Working with matrices

Practice 1.0) Let's create some matrices
```{r eval=FALSE}

myvec = c(1:12)

dim(myvec) = c(3,4)

myvec
     
mymat = matrix( c(7, 4, 2, 4, 7, 2), nrow=3, ncol=2) 

mymat
```

***

1.1) You can influence how the matrix is constructed

```{r eval=FALSE}

mymat = matrix( c(7, 4, 2, 4, 7, 2), nrow=3, ncol=2, byrow=TRUE)
mymat

```

***

1.2) You can name the rows and columns of a matrix according to a pattern

```{r eval=FALSE}
set.seed(123)

X = matrix(rpois(20,1.5),nrow=4)

X    

rownames(X) = paste("Trial",1:nrow(X),sep=".")

X
       
colnames(X) = paste("P",1:ncol(X),sep=".")

X

```

***

1.2) You can also name columns something more specific

```{r eval=FALSE}

drug.names = c("aspirin","paracetamol","nurofen","hedex","placebo")

colnames(X) = drug.names

X
        
X['Trial.1',]   # Gets all columns for Trial #1

# Get's the nurofen column for Trial.1

X['Trial.1','nurofen']

X[,'nurofen']                     # Get all Trials for nurofen

X[,'nurofen',drop=FALSE]   # Preserves matrix structure if desired

```

***

1.3) Indexing into Matrices are usually done numerically

```{r eval=FALSE}

set.seed(123)

X = matrix(rpois(9,1.5),nrow=3)

X

X[1,1]     # First row, First Column

X[2,2]     # Second row, Second Column

X[3,3]     # Third row, Third column

diag(X)    

X

X[1:2,1]     # Gets First and second rows and the first column

X[1:2,2]     # Gets First and second rows and the second column

X[1:2,]      # Gets First and second rows and ALL columns

```

***

1.4) 

```{r eval=FALSE}

X[1:4]

X >= 2

X[X >= 2]   # Returns which values are greater or equal to 2

which(X >= 2)   # Returns which elements are greater or equal to 2

X %% 2 == 0          # Returns a logical matrix 
      
X[ X %% 2 == 0 ]   # Finds the actual element values
    
X[X %% 2 == 0] = 99  # Set all even numbers to 99

```

***

1.5) We can use the row and column functions to find elements that satisfy specific conditions. This approach takes some getting used to it but its better than using a for and if loop combination.

```{r eval=FALSE}
X
     
row(X)
     
col(X)      

row(X) == col(X)

X[row(X) == col(X)]

X[row(X) == col(X)] = 0  # Put zeroes in the diagonal

```

***

1.6) We can use the cbind and rbind functions to add columns and rows to a matrix. This is necessary when we have several matrices we would like to combine.

```{r eval=FALSE}
set.seed(123) 

X = matrix(rpois(9,1.5),nrow=3)

colnames(X) = c("aspirin","paracetamol","nurofen")

rownames(X) = paste("Trial",1:3,sep=".")

rbind(X,Trial.4=c(4,7,5))

X
    
rowSums(X)

cbind(X,rowsums = rowSums(X))

```

***

1.7) Let's use some functions on a matrix to show that we can extract some meaningful information

```{r eval=FALSE}

set.seed(123) 

X = matrix(rpois(9,1.5),nrow=3)

colnames(X) = c("aspirin","paracetamol","nurofen")

rownames(X) = paste("Trial",1:3,sep=".")

X
        
mean(X[,3])  # Mean of the 3rd column

var(X[3,]) # Variance of the 3rd row

X

rowSums(X)

colSums(X)

rowMeans(X)
  
colMeans(X)
  
colMeans(X)[3]

```

***

1.8) We can also do some cool linear algebra operations

```{r eval=FALSE}

A = matrix(c(1,3,2,2,8,9),3,2)

t(A)

A

B = matrix(c(5,8,4,2),2,2)

A %*% B

A = matrix(1:4,2,2)

A
    
B = solve(A)

B

A %*% B            # We get the identity matrix

A

b = c(7,10)

x = solve(A) %*% b

x

# Slide 25

B

diag(B)             # Fetches the diagonal


diag(c(1,2,3))      # Creates a matrix with 1,2,3 on the diagonal
     

diag(1,4)           # Creates a 4 x 4 Indentity matrix

```

***

1.9) Get used to the apply function. It does some cool things

```{r eval=FALSE}
set.seed(123) 

X = matrix(rpois(9,1.5),nrow=3)

colnames(X) = c("aspirin","paracetamol","nurofen")

rownames(X) = paste("Trial",1:3,sep=".")

apply(X,2,function(var) var/sum(var))

apply(X,1,function(var) var/sum(var))

# Find the distance between each element and the mean of its respective column

apply(X, 2, function(x) x - mean(x))

```

## Working with Factors

Problem 2.0) Factors

```{r eval=FALSE}

transvec = c(1,1,1,0,0,0,0,0,0,0,0,0,0,0,0,0,0,1,1,1,0,0,0,0,0,1,1,1,1,1,1,1)

table(transvec)    # Count 'em up. Which are Auto and Manual ?

mytransfac = factor(transvec, levels = c(0,1), labels = c("Auto","Man") )

table(mytransfac)

levels(mytransfac)

mytransfac

library(lattice)

xyplot(mpg~wt | mytransfac, mtcars, main="MPG vs Weight - Auto and Manual Transmissions")

bwplot(~mpg|mytransfac, mtcars, main="MPG - Auto and Manual Transmissions",layout=c(1,2))

```

***

2.1) We can cut up continuous quantities into discrete categories and, ultimately, factors

```{r eval=FALSE}
mympg = c(21,21,22.8,21.4,18.7,18.1,14.3,24.4,22.8,19.2,17.8,16.4,17.3,15.2,10.4,
        10.4,14.7,32.4,30.4,33.9,21.5,15.5,15.2,13.3,19.2,27.3,26,30.4,15.8,19.7,15,21.4)

tapply(mympg,mytransfac,mean)
    
# Slide 38 

cut(0:10,breaks=4)

table(cut(0:10,breaks=4))

# Slide 39

my.cut = cut(0:10,breaks=4,labels=c("Q1","Q2","Q3","Q4")) 

table(my.cut)

```

***

2.2) More on cutting continuous quantities into factors

```{r eval=FALSE}

quantile(0:10)
  
table(cut(0:10,breaks=quantile(0:10),include.lowest=TRUE))

set.seed(123)

exam.score = runif(25,50,100)

cut(exam.score,breaks=c(50,60,70,80,90,100))

cut(exam.score,breaks=c(50,60,70,80,90,100),labels=c("F","D","C","B","A"))
 
my.table = table(cut(exam.score,breaks=c(50,60,70,80,90,100),labels=c("F","D","C","B","A")))

barchart(my.table,main="Grade BarChart",col=terrain.colors(5))

cut(exam.score,breaks=c(50,60,70,80,90,100))
 
cut(exam.score,breaks=c(50,60,70,80,90,100),right=F)

```

***

2.3) If you don't use cut then you have to write for and if loops to do the work

```{r eval=FALSE}
exam.score = runif(25,50,100)

acount = 0
bcount = 0
ccount = 0
dcount = 0
fcount = 0
exam.score = runif(25,50,100)
for (ii in 1:length(exam.score)) {
  
 if (exam.score[ii] < 60) {fcount = fcount + 1} else
   if ((exam.score[ii] >= 60) & (exam.score[ii] < 70)) {dcount = dcount + 1} else
     if ((exam.score[ii] >= 70) & (exam.score[ii] < 80)) {ccount = ccount +1} else
       if ((exam.score[ii] >= 80) & (exam.score[ii] < 90)) {bcount = bcount +1} else
          if ((exam.score[ii] >= 90) & (exam.score[ii] <= 100)) {acount = acount +1}
}

cat("acount bcount ccount dcount fcount \n")

cat(acount,bcount,ccount,dcount,fcount,"\n")

```

***

2.4) We can also impose an order on the factors we create:

```{r eval=FALSE}

mons =c("Jan","Feb","Mar","Apr","May","Jun","Jan","Feb","May","Jun", "Apr","Mar")

my.fact.mons = factor(mons)

my.fact.mons[1] < my.fact.mons[2]

levels(my.fact.mons)

my.fact.mons = factor(mons, labels=c("Jan","Feb","Mar","Apr","May","Jun"),ordered=TRUE)

my.fact.mons

my.fact.mons[1] < my.fact.mons[2]

table(my.fact.mons)

levels(my.fact.mons)                       # This is what we want !

```

***
