BIOS545 Spring 2015 Week 4 Lab 1:Functions
============================================================
Pittard wsp@emory.edu
***

## Functions

PRACTICE 1.0) VIEWING THE SOURCE CODE OF SOME R FUNCTIONS IS POSSIBLE:
```{r eval=TRUE}

ls

```
***

PRACTICE 1.1) SOME OF THESE FUNCTIONS REFER TO "METHODS". WE HAVE TO WORK A LITTLE HARDER TO SEE THE SOURCE CODE FOR THEM:

```{r eval=TRUE}

t.test

methods(t.test)

getAnywhere(t.test.default)
```
***

PRACTICE 1.2) THERE ARE SOME FUNCTIONS THAT CAN SHOW US WHAT THE ARGUMENTS ARE FOR A FUNCTION AS WELL AS WHAT AN EXAMPLE MIGHT LOOK LIKE:

```{r eval=TRUE}

args(ls)

args(mean)

example(mean)

```
***

PRACTICE 1.3) DECLARING A FUNCTION IS EASY:

```{r eval=FALSE}

my.func <- function(somenum) {
  my.return.val <- sqrt(somenum)
  return(my.return.val)
}

# We could do it all with one line but that isn't good for readability:

my.func <- function(somenum) {return(sqrt(somenum))}

```
***

PRACTICE 1.4) HERE IS AN EXAMPLE FOR THE pythagorean theorem
```{r eval=TRUE}

pythag <- function(a,b) {
 c <- sqrt(a^2 + b^2)
 return(c)
}

pythag(4,5)

x <- 4
y <- 5

pythag(x,y)

pythag(a=4, b=5)

```
***

PRACTICE 1.5) WE CAN RETURN A LIST OF INFORMATION
```{r eval=FALSE}
pythag <- function(a,b) {
    c <- sqrt(a^2 + b^2)
    myreturnlist <- list(hypoteneuse = c, sidea = a, sideb = b) 
    return(myreturnlist)
}

pythag(3,4)  

pythag(3,4)$hypoteneuse

```
***

PRACTICE 2.0) WE NEED TO VERIFY WHAT THE USER IS PASSING TO US IS VALID INPUT. IF WE DON'T THEN WE CAN GET UNEXPECTED RESULTS
```{r eval=TRUE}
pythag <- function(a,b) {
    c = sqrt(a^2 + b^2)
    myreturnlist <- list(hypoteneuse = c, sidea = a, sideb = b) 
    return(myreturnlist)
}

pythag(3,4)

# pythag(3,"a")  # This will bomb

# pythag()

# pythag(3,)

```
***

PRACTICE 2.1) WE NEED TO VERIFY WHAT THE USER IS PASSING TO US IS VALID INPUT. IF WE DON'T THEN WE CAN GET UNEXPECTED RESULTS
```{r eval=TRUE}
pythag <- function(a = 4, b = 5) {
     if (!is.numeric(a) | !is.numeric(b)) {
   		stop("I need real values to make this work")
     }
     c <- sqrt(a^2 + b^2)
     myreturnlist <- list(hypoteneuse = c, sidea = a, sideb = b) 
     return(myreturnlist)
}

# pythag(3,"5")  # These will bomb - try it out

# pythag("3",5)

```
***

PRACTICE 2.2) WE NEED TO VERIFY WHAT THE USER IS PASSING TO US IS VALID INPUT. IF WE DON'T THEN WE CAN GET UNEXPECTED RESULTS
```{r eval=TRUE}
pythag <- function(a = 4, b = 5) {
     if (!is.numeric(a) | !is.numeric(b)) {
   		stop("I need real values to make this work")
     }
     if (a <=0 | b <= 0) {
 			stop("Arguments need to be positive")
     }
     c <- sqrt(a^2 + b^2)
     myreturnlist <- list(hypoteneuse = c, sidea = a, sideb = b) 

     return(myreturnlist)

} # End Function

# pythag(-3,3)   # This will bomb

pythag(3,3)

```
***

PRACTICE 3.0) FUNCTIONS NEED TO BE ABLE TO ACCOMMODATE A RANGE OF DATA TYPES. RARELY DO WE DEAL WITH JUST SINGLE VALUES.

```{r eval=TRUE}
is.odd <- function(someval) {
  retvec <- vector()
  for (ii in 1:length(someval)) {
    if (someval[ii] %% 2 != 0) {
        retvec[ii] = TRUE
    } else {
        retvec[ii] = FALSE
    }
  }
  return(retvec)

}  # End function

is.odd(3)

numbers <- c(9,9,4,4,6,10,7,18,2,10)

is.odd(numbers)

numbers[is.odd(numbers)]   # Very useful

```
***

PRACTICE 3.1) LET'S WRITE SOME FUNCTIONS TO PROCESS VECTORS AND AT THE SAME LEARN HOW TO STRUCTURE THE CODE INTERNAL TO FUNCTIONS

```{r eval=TRUE}
#HERE WE HAVE SOME CODE THAT WOULD PROBABLY BE MORE APPROPRIATE FOR A FUNCTION

y <- vector()
x <- seq(-3,3)      
for (ii in 1:length(x)) {
  y[ii] <- (x[ii])^2 
}

length(x)

plot(x,y,main="Super Cool Data Plot",type="l")

```
***

PRACTICE 3.2) LET'S CREATE A FUNCTION TO DO THE ABOVE:
```{r eval=TRUE}

myplotter <- function(xvals, mfunc, plotcolor="blue") {

# Function to print y = x^2
# Input: xvalues
# Output: A plot and the xvals and yvals used to make that plot

   yvals = vector()
   for (ii in 1:length(xvals)) {
     yvals[ii] <- mfunc(xvals[ii])
   }
  
   plot(xvals, yvals, main="Super Cool Data Plot",type="l",col=plotcolor)
   retlist <- list(x=xvals, y=yvals)
   return(retlist)
}

xvals <- seq(-3,3,0.5)

myplotter(xvals, cos, plotcolor="red")

```
***

PRACTICE 3.3) WE'LL NOW CREATE A FUNCTION THAT FINDS THE MINIMUM OF A VECTOR:
```{r eval=TRUE}
mymin <- function(somevector) {

# Function to find the minimum value in a vector
# Input: A numeric vector
# Output: A single value that represents the minimum

  mymin <- somevector[1] # Set the minimum to an arbitrary value

# Now loop through the entire vector. If we find a value less than 
# mymin then we set mymin to be that value.

  for (ii in 1:length(somevector)) {
    if (somevector[ii] < mymin) {
      mymin <- somevector[ii]
    }
  }
  return(mymin)
}

set.seed(123)

testvec <- rnorm(10000)

mymin(testvec)

min(testvec)  # Matches the built in R function

```
***

PRACTICE 3.4) WE'LL NOW EXTEND THIS TO ACCOMMODATE OPTIONS FOR FINDING A MAX OR MIN
```{r eval=TRUE}
myextreme <- function(somevector, action="min") {

  if (action == "min") {
     myval <- somevector[1] # Set the minimum to an arbitrary value

     for (ii in 1:length(somevector)) {
       if (somevector[ii] < myval) {
         myval <- somevector[ii]
       }
     }    # End for

  } else {   # If action is not "min" then we assume the "max" is wanted

     myval <- somevector[1] # Set the minimum to an arbitrary value

     for (ii in 1:length(somevector)) {
       if (somevector[ii] > myval) {
         myval <- somevector[ii]
       }
     }            # End for
  }               # End If  
  return(myval)
}

myextreme(testvec,"min")

myextreme(testvec,"max")

min(testvec)

max(testvec)

```
***


PRACTICE 4.0) WE CAN WRITE FUNCTIONS THAT HELP US PROCESS "SPLIT" DATAFRAMES:
```{r eval=TRUE}
myfunc <- function(somedf, somefac) {
  
# Function to split a data frame by a given factor
# Input: A data frame, a factor
# Output: A list containing a count of records in each group

  retlist <- list()    # Empty list to return group record count 
  mysplits <- split(somedf,somefac)  # Split the data frame by somefac
  
  for (ii in 1:length(mysplits)) {  # loop through the splits
    retlist[[ii]] <- nrow(mysplits[[ii]]) 
  }
  names(retlist) <- names(mysplits)
  return(retlist)
}

myfunc(mtcars,mtcars$cyl)

```
***


PRACTTICE 4.1) WE CAN WRITE FUNCTIONS THAT HELP US PROCESS MATRICES:
```{r eval=TRUE}
mtcenter <- function(somemat) {

# Input: A matrix to center
# Output: A matrix that is centered

    retmat = rep(0, length(somemat)) # Recipe to initialize a 
    dim(retmat) = dim(somemat)       # matrix the same size as
                                     # another filled with 0

    for (col in 1:ncol(somemat)) {
      for (row in 1:nrow(somemat)) {
        retmat[row, col] <- somemat[row, col] - mean(somemat[,col])
      }
    }

    return(retmat)
}

set.seed(123)

mymat <- matrix(round(rnorm(6),2),3,2)

mtcenter(mymat)
```
***

PRACTICE 4.2) IN R WE USE LOTS OF ANONYMOUS FUNCTIONS TO DO WORK ALTHOUGH THEY AREN'T ALWAYS EASY TO UNDERSTAND:
```{r eval=TRUE}
my.mat <- as.matrix(mtcars[,c(1,3:6)])

head(my.mat)

apply(my.mat,2, mean)

my.mat <- as.matrix(mtcars[,c(1,3:6)])

head(my.mat)

apply(my.mat,2, function(x) {c(mean=mean(x),sd=sd(x),range=range(x))})

apply(my.mat,2, function(x) {
                     c(mean=mean(x),
                     sd=sd(x),
                     range=range(x))
                  })
```
***
